%&cheri_compartments_preamble
\endofdump

\begin{document}
\maketitle

\section{Introduction}

CHERI capabilities implicitly portion a program into \emph{compartments}, but
there are two fundamentally different flavours of creating CHERI compartments
-- both of which can be used within a single process -- and several
sub-flavours of each. Our aim in this draft paper is to outline the different
forms of compartments that CHERI enables.


\section{Background}

We assume high-level familiarity with capabilities (an approachable historical overview
is~\cite{levy84capability}, which may usefully be augmented by more recent work
such as \cite{miller06robust}),
CHERI generally (see e.g.~this introduction~\cite{watson20cheriintroduction}),
the CHERI ABI~\cite{brooks19cheriabi},
and CHERI C (the dialect of C that CHERI defines)~\cite{watson20chericprogramming}.

Because CHERI has been developed over a number of years, and is explained over
a variety of documentation and papers, some concepts have more than one name.
We have found that some of those names make it difficult to talk, and sometimes
reason about, capability usage.
Unfortunately, we can think of no better solution to this than to propose our
own terminology.
\jacob{That sticks out as being a bit too informal in the context. Suggestion:
"To avoid ambiguity, we will explicitly define key terms that we use."}

We use the term \textit{CHERI} to refer to the `abstract capability machine'
that software can observe: that is, the combination of a capability hardware
instruction set, an ABI, and a user-facing library that exposes capability-related functions.
We refer to specific hardware implementations by their name
(e.g.~Arm's `Morello', or `CHERI RISC-V').

\jacob{I'm not sure if they'd be considered "immutable". At least, no more than
any other value in an architecture with a mostly constructive ISA. \emph{Sealed}
capabilities are immutable (except for unsealing and de-authenticating
operations) and that distinction may be useful later.}
CHERI capabilities are immutable and thus a new \emph{child}
capability must be derived from one or more \emph{parent} capabilities.
We refer to a capability as \emph{authentic} (what CHERI calls
`tagged' or `valid') if it has been derived from its parents according
to CHERI's rules, or \emph{inauthentic} (what CHERI calls `untagged' or `invalid')
otherwise. A capability consists of an \emph{address} in one or more dimensions
(usually memory) and a set of \emph{permissions}. Amongst the permissions are
\emph{bounds} -- the region of space in which an authentic capability is allowed to
operate.
A capability's bounds are from a \emph{low} (inclusive) to
a \emph{high} (exclusive) address and we refer to `a bound of $x$ bytes' to mean
a capability where $\textit{high}-\textit{low}=x$.
If a capability's address is contained within its bounds we refer to the capability as
a whole as \emph{in-bounds} or \emph{out-of-bounds} otherwise
(see~\cite{woodruff19chericoncentrate} for an explanation of why an authentic
capability might have an out-of-bounds address). Other permissions include
boolean flags such as whether a capability can read/write to memory addresses
within its bounds. Such permissions also define the dimensions to which
the address refers: a capability with a \texttt{LOAD} permission
inherently addresses memory, for example.
\jacob{I'm conflicted here between offering general definitions, which may be
hard to follow, or offering memory capabilities as an example. I've changed it
to make it more general, but was that the right decision? Feel free to revert.}

As well as capabilities (which on Morello and CHERI RISC-V, for example, are
implemented as double-width addresses), CHERI also allows `traditional' single-width addresses
to be used. Conventionally, a program which uses both traditional addresses and
capabilities is said to be operating in \emph{hybrid} mode while a program
which uses only capabilities is in \emph{pure capability} (or \emph{purecap})
mode. However, these terms are tied to specific C implementations and are often
insufficient when discussing other language models, or language-agnostic
compartments. We will avoid unqualified use of these terms.
\jacob{I just added that statement, based on our recent discussions. I realise
that the next section immediately talks about "pure capability CHERI program"
without (much) qualification. I think the general structure (and especially
Figure~\ref{fig:reachability}) is right, but we might need to check that it
flows correctly and isn't misleading, perhaps adding some detail to qualify what
"pure capability compartmentalisation" means to us.}


\section{Pure capability compartmentalisation}

\begin{figure}
\begin{center}
\begin{tikzpicture}[
  align=center,
  node distance=2.8cm and 4cm,
  every by arrow/.style={*->},
  every edge/.style={draw,->,>=stealth',
  auto,
  semithick
}]
\node[state, shape=rectangle] (B1) {};
\node[state, shape=rectangle, right of=B1] (B2) {};
\node[state, shape=rectangle, right of=B2] (B3) {};
\node[state, shape=rectangle, right of=B3] (B4) {};
\node[state, shape=rectangle, right of=B4] (B5) {};
\path[->] (B1) edge[above] node{C1} (B2);
\path[->] (B2) edge[bend left, above] node{C2} (B3);
\path[->] (B3) edge[bend left, below] node{C3} (B2);
\path[->] (B3) edge[above] node{C4} (B4);
\path[->] (B4) edge[above] node{C5} (B5);
\end{tikzpicture}

\vspace{10pt}

\begin{tabular}{ll}
\toprule
Root set & Reachable set \\
\midrule
\{C1\} & \{C1, C2, C3, C4, C5\} \\
\{C2\} & \{C2, C3, C4, C5\} \\
\{C3\} & \{C2, C3, C4, C5\} \\
\{C4\} & \{C4, C5\} \\
\{C5\} & \{C5\} \\
\bottomrule
\end{tabular}

\caption{Pure capabilities form a graph: the root set of capabilities a program
has access to determine the full set of reachable capabilities the
program can reach. The diagram above represents a simplification of the
running state of a pure capability program: each square constitutes a
block of memory containing zero or more capabilities; an outgoing edge
from a block signifies that it stores a capability pointing to another
block; each capability has the \texttt{LOAD\_CAP} permission set so
that it can load capabilities from the block of memory points to; and
each capability has its bounds set to cover
exactly one other block of memory. The table then shows the graph of
reachable capabilities if, at a specific point in time, the root set of
capabilities (i.e.~those capabilities the program can immediately
access because they are stored in e.g.~registers) contains exactly one
element.}
\end{center}
\label{fig:reachability}
\end{figure}


Pure capability CHERI programs implicitly or explicitly use capabilities to
define compartments. Figure~\ref{fig:reachability} shows an example of one
particular flavour of compartment, but there are others. In this Section we
show examples of different, overlapping, flavours of pure capability
compartments.


\subsection{Memory compartmentalisation}

\laurie{LOAD\_CAP, I think, implicitly defines this. entirely dynamic.}
\jacob{If my intuition about what "memory compartmentalisation" is is correct,
then yes. We do, however, have to be careful not to inadvertently discount other
capabilities; a capability that grants LOAD (or something else) is still part of
the compartment, even in the absence of LOAD\_CAP. It may be able to read a
cyptographic key or some other non-capability secret, for example.}


\subsection{Link-time compartmentalisation}

\laurie{What Hesham calls `ELF/linkage-based or library-based compartmentalization'}
\jacob{This implies something about where the boundaries are set, but is the
actual compartmentalisation different from "memory compartmentalisation"?
(That's not rhetorical; I wonder if the GOT needs to be handled differently, for
example.)}

\section{Hybrid compartmentalisation}

\subsection{Hybrid CHERI data compartmentalisation}
\al{Just a small note, that, also related to your other comment, I don't have a
feel for the difference between data and code compartmentalisation.}
\laurie{at a simple level my feeling is that: `data compartmentalisation' ==
`restrict the DDC' and `code compartmentalisation' == `restrict the PCC'. you
can do either separately or both together (i.e.~there are three modes in this
view of the world)}
\al{With a few days thinking, I get the difference now. I do think what I'm
outlining here is join data/code compertmentalization.}

\laurie{i assert that this section is
probably better called `Hybrid data compartmentalisation', because we're not
compartmentalising the PCC, and i think the techniques are trivially adaptable
to RISC-V (or MIPS or ...).}
\al{As long as certain features/instructions are available in those
architectures, but I am unfamiliar with non-Morello CHERI ISAs to be able to
say. For instance, immediately sealing with a literal (as we do for the indirect
capability, to be used with `lpb` instructions, I believe is not a CHERI thing,
but a Morello one --- although I might be wrong).}\laurie{aha, let's ask jacob
or otherwise try to work out if this is morello only or not. i reckon that
it's probably a generic cheri feature to some extent, but there's probably
not a nice api for it, so you have to hack it on each platform. i might be wrong!}
\jacob{CHERIv8 (TR 951) has CSealEntry, which looks similar, though it can only
use one seal type. I'm not sure if that's an important limitation. Even without
CSealEntry, you could do this by making an appropriate sealing capability
available to every compartment, so the general technique is availble to all
CHERI implementations.}

\laurie{one thing this paragraph made me realise is that if the user creates
$n$ data compartments at run-time then we actually have $n+1$ compartments as
the switcher is, itself, implicitly a compartment.}
\al{Implicitly, yes. There must be a mechanism to perform the mechanics required
for the transition, and in order to make that mechanism opaque to compartments,
it can either (a) function at a full privileged level, or (b) be within its own
small compartment. You can argue as well that the whole system (with unbound
PCC/DDC) is a compartment.}
\jacob{Perhaps "at least $n$ compartments" is appropriate: In our discussion
last week, we decided that removing a capability derives a new, lesser
compartment from the current one. If the process of scrubbing private
capabilities (to prepare for a compartment switch) requires several steps, then
multiple (temporary) compartments will be created. On Morello, each capability
register will need to be individually cleared, so there will usually be several
steps! Since they are all subsets of the starting point, this distinction is
probably not important. However, sometimes it \emph{is} important, and I have
struggled to robustly define criteria for determining when that is the case.}

Supposing compartmentalisation is performed between two unprivileged
compartments via a privileged entity, it makes sense for an attacker to focus
their efforts attacking that privileged compartment switcher, to perhaps gain
access to the whole system. One possible approach to tighten security is to
minimize the functionality of such a switcher: by identifying what it means to
perform a transition from one compartment to another, we can prepare a switching
mechanism that only performs the required instructions, with minimal interaction
with the rest of the environment, and is thus less likely to be misused.

To exemplify, we model a simple system with two compartments (and will be
abstracting away some of the finer details, such as memory management). Suppose
we have two compartments, named \(f\) and \(g\). To further simplify this model,
we shall assume each compartment is comprised of a single function of interest,
and a compartment transition
\jacob{Aren't there two compartment transitions involved in a call like this?}
involves executing that particular function, then
returning. We shall further assume that each compartment has its own memory
region, which serves as both stack and heap (again, specific details of memory
allocation will be omitted), disjoint with memory regions of the other
compartment.

(TODO: explain DDC/PCC)
With the above model, a compartment transition would involve:
\begin{itemize}
\item identify the target compartment to be transitioned to, effectively
executing the corresponding compartment function;
%
\laurie{obvious question: how does one compartment say `switch to another
compartment'? perhaps more precisely: what's the mechnanism for identifying a
compartment and do we guarantee that it's unforgeable? do we hand out sealed
capabilities to represent compartments or ... ?}
\al{In the current implementation, as presented here, we simplify to have each
compartment have a single entry-point (i.e., a executable function covered by
the PCC corresponding to that compartment),  and then transitioning to a
compartment would involve jumping to execute that function. In a ``proper''
implementation, I imagine you would have essentially a map from compartment to
functions, more specifically a compartment identifier (could be hash), and a
list of capabilities to be used for jumping to functions that the respective
compartments wants to expose, in a library-like fashion. So yes, at compartment
creation time, a way would be to create capabilities for functions you want to
expose, and either copy those to all compartments, or perhaps hash them somehow,
store the capabilities in the switcher for practical use, but each compartment
would know what hash to use when it want to call a specific function from
another compartment.}
\laurie{let me try asking a variant: if i have $n$ compartments in my
system and I want to switch to compartment $C_{n - 1}$, I need to have some
sort of identifier that says ``switch to $C_{n - 1}$ and execute some
function''. My question is: what is the identifier? If it's a hash I can
forge it. That suggests the ID should probably be a sealed capability?}
\al{The simple answer is, at its core, you will execute a function in a
container (therefore, transition), via a provided capability. The long answer
is, that when we create a container, we would need to know which functions that
we place within that container should be callable from others, and create
appropriate capabilities. So, to call a function would be needing the DDC of the
target container (already known ahead of time, as presumably that shouldn't
change), and then the respective capability for the desired function, which
would be installed as the PCC on transition. So what I was suggesting above is
store all these PCC-capabilities (sealed) in a hashmap in the switcher, then use
the hash to identify what capability the switcher needs to select for the PCC
(thinking about it, the hash can just be the address of the target function). So
you won't be able to forge it, as the hash/address/whatever won't give you
access to the actual capability you will need to execute (which, in turn, would
also be sealed to only be used via branch instructions).}
\laurie{ok, so you're proposing here combined data and code
compartmentalisation. i think that code compartmentalisation isn't super
useful for us at the moment (or, at least, until we work with JITed code)
and it would make our lives to concentrate just on data
compartmentalisation. that would allow us to explore the use case of
``let's create multiple instances of the Lua interpreter within one Unix
process and compartmentalise each'' -- by definition, we can't do code
compartmentalisation in this case.}
\al{As I said during the call, having both should not preclude being able to do
the lua example, and I think I'm close enough to finalizing this approach than
having to redesign stuff. We can re-evaluate once it's done, or it's taking
longer than the few days I anticipate, or we get into practical problems trying
to run the lua example we want.}
\jacob{I agree with the conclusion arrived at above -- though I think there
probably are uses for code compartmentalisation outside of JITs -- but I think
that today's prototype, with a single function per compartment, actually scales
to multiple functions quite nicely if we implement it as a different compartment
with the same bounds (data and code), but a different entry point. Another
point, though: if the caller compartment has direct access to the sentry, can
they misuse it? That is, if the callee relies on the compartment switcher to
clean anything up, then it must be sure that it can only be called from the switcher.
In such an implementation, we would indeed need some other, abstract sealed
capability. However, that's an implementation detail.}
%
\item install a capability in the \texttt{DDC} which covers the memory region
corresponding to the target compartment;
\jacob{Where does that come from? If the caller has it (e.g. in a register or in
its reachable memory) then it's \emph{already} part of the compartment, and this
initial model doesn't describe a compartment switch at all.}
\item install a capability pointing to the target function in the \texttt{PCC},
and start execution;
%
\laurie{ah! i actually wonder if we should do a second
section called `hybrid memory + code compartmentalisation'? that way we can
keep this first section simple. i suppose one can also have `hybrid code
compartmentalisation' on its own too, but i can't immediately think of a
great use case for that.}
\jacob{If we're switching data compartment, we \emph{must} also set the PCC to
one of a set of predefined entry points. If we don't enforce that, then code
could switch data compartments without also transferring execution control,
which effectively gives compromised code access to all data compartments it can
switch to (transitively). As a result, this bullet point is important. This
subtlety is probably worth describing somewhere. Do we consider that to be "code
compartmentalisation", or just a form of forward-edge CFI? I don't know.}
%
\item upon returning from executing the target function, reinstall the
\texttt{DDC} and \texttt{PCC} corresponding to the source compartment memory region.
\end{itemize}
The above process involves some level of trust between compartments, as they
require access to capabilities defining memory accessible by the other
compartment
%
\laurie{i think this might be related to my question above about `how we identify a compartment'?}
\al{And this is to add to the above ``obvious question'', at compartment
initialization time, we would create the capabilities defining that
compartment's executable and memory regions, to which other compartments won't
have access to (due to DDC bounding), and of which more can't be created (unless
the compartment itself decides to create some and leak).}
%
, as well as executable code within that compartment. Having access
to these capabilities means lack of CHERI security guarantees over these ares in
memory. Thus, we can strengthen our model by suggesting \emph{mutual distrust}
between compartments themselves. This involves adding a level of indirection in
the above process. We define a concrete ``transition state'', comprising of both
executable code involving setting the state for compartment transitions (i.e.,
installing capabilities with correct bounds in \texttt{PCC} and \texttt{DDC}
registers), as well as a memory region containing compartment information (the
aforementioned capabilities defining compartments' allocated memory regions, and
visible executable code).

Let's update the process above, by adding a setup phase to create required
capabilities ahead of time, and to introduce indirection during the compartment
transition function. We require the following capabilities be constructed ahead
of time, in a privileged environment:
\begin{itemize}
\item for each compartment, a capability defining the accessible memory region
of that compartment;
\item for each compartment, a capability defining the executable part of that
compartment which should be callable by other compartments;
\item a capability defining a separate memory region to hold the above
capabilities;
\item a capability covering the code which will perform the transition;
\item a sealed (discussed shortly) capability that points to the memory address
where the previous two capabilities are stored, consecutively, and in that
order.
\jacob{The order is important for Morello but that's an implementation
detail. It's really just important that they're accessible only as a pair.}
\end{itemize}
The first two types of capabilities will be stored under the memory region
allocated for compartment transitioning. The following two capabilities will
similarly be stored in the same memory region, but available to the compartments
indirectly by the final capability (which shall be discussed in more detail
shortly). Copies of the final capability are provided to each compartment, to be
stored within their respective memory regions, and to facilitate compartment
transitioning.

With the above setup in place, we can observe that
\begin{enumerate}
\item during compartment transitioning, we have access to data defining each
compartment's visible executable code and accessible memory,
\item and compartment transitioning is done within its own, separate scope, and
does not give full permissions over the whole system.
\end{enumerate}
However, it is important to note that this transition state does not have
privileged execution. It is indeed the case that were a malign actor gain
control of the memory region where comaprtment information is stored, it would
be able to gain control of all compartments using that specific transition code.
However, it would not gain control of any other part of the whole system.

\jacob{Is this too implementation-specific? Do we need an "in our prototype"
qualifier somewhere?}
The actual process by which the switching code is invoked is via the Morello
\texttt{ldpblr} instruction\cite{...}, which stands for ``Load Pair Branch with
Link to Register''. Given a register containing a capability, the instructions
dereferences the memory address indicated by the value of the capability, reads
two consecutive capabilities stored at that memory address, installs the first
capability in \texttt{C29}, and performs a \texttt{blr} with the second
(meaning installing the second in the \texttt{PCC}, and updating the
\texttt{CLR} to the following instruction, to facilitate returning from the
called function). This level of indirection allows a compartment to essentially
change the state defined by the environment, by setting both \texttt{C29} and
\texttt{PCC} at once, without actually having access to them both as they would
potentially reside outside memory defined as accessibly by the compartment's own
\texttt{DDC} capability. A hybrid callee can then install \texttt{C29} into its own
\texttt{DDC}; a purecap callee may use it to restore its capability stack
pointer (\texttt{csp}) and other general-purpose registers. Further, we can
\emph{seal} the capability providing the indirect memory address of the
switcher's capabilities with a Morello-specific object type of \texttt{lpb},
indicating the capability should only be used as part of \texttt{lpb}-like
instructions (which include \texttt{ldpblr}). This, however, requires the
indirect memory capability be installed in register \texttt{C29}, as the
unsealing operation is only performed if that is the source register provided to
\texttt{ldpblr}.
\jacob{I corrected this to \texttt{C29}. Readers may point out that other
registers may be specified, but only \texttt{C29} allows the automatic unsealing
behaviour, which gives the callee an important guarantee. (We probably need to
examine that guarantee somewhere.)}






\bibliographystyle{plain}
\bibliography{bib}

\end{document}
